---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# odkmerge

<!-- badges: start -->
[![R-CMD-check](https://github.com/kcham193/odkmerge/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kcham193/odkmerge/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/odkmerge)](https://CRAN.R-project.org/package=odkmerge)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)
[![Codecov test coverage](https://codecov.io/gh/kcham193/odkmerge/branch/master/graph/badge.svg)](https://app.codecov.io/gh/kcham193/odkmerge?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**Merge and flatten ODK Central and KoboToolbox repeat-group exports in one line of R.**

ODK Central and KoboToolbox both export repeat-group data as separate sheets
in an `.xlsx` file. Joining them back together manually is tedious and
error-prone. `odkmerge` detects the sheet structure automatically and produces
flat, analysis-ready tibbles — for any form, regardless of complexity.

Both export conventions are auto-detected:

| | Row identifier | Parent reference |
|---|---|---|
| **KoboToolbox** | `_index` | `_parent_index` |
| **ODK Central** | `KEY`    | `PARENT_KEY`    |

System columns (`_submission_*` on Kobo; `SubmissionDate`, `SubmitterID`,
`Status`, etc. on Central) are stripped by default so the merged data is
analysis-ready.

## How It Works

```
┌─────────────────────┐     ┌─────────────────────┐
│   survey (parent)   │     │   species (repeat)   │
│  _index │ plot_id   │     │ _index│_parent_index │
│    1    │  P01      │     │   1   │      1       │
│    2    │  P02      │     │   2   │      1       │
│   ...   │  ...      │     │   3   │      2       │
└─────────────────────┘     └─────────────────────┘
            │                          │
            └──────── odkmerge ────────┘
                          │
                          ▼
        ┌──────────────────────────────────────┐
        │         master dataset               │
        │ _parent_index │ plot_id │ species    │
        │      1        │  P01    │ Acacia     │
        │      1        │  P01    │ Combretum  │
        │      2        │  P02    │ Themeda    │
        └──────────────────────────────────────┘
```

## Installation

```{r install, eval = FALSE}
# GitHub (development version):
# install.packages("devtools")
devtools::install_github("kcham193/odkmerge")

# CRAN (once published):
install.packages("odkmerge")
```

## Quick Example

```{r example, message = FALSE}
library(odkmerge)

path   <- system.file("extdata", "simple_survey.xlsx", package = "odkmerge")
master <- odk_merge(path, verbose = FALSE)
head(master)
```

For multi-repeat or nested forms, `odk_merge()` returns a named list of tibbles
— one per repeat group.

## Step-by-step control

```{r workflow, eval = FALSE}
# 1. Read all sheets
sheets <- read_odk_export("my_kobo_export.xlsx")

# 2. Inspect the structure
detect_structure(sheets)

# 3. Build the flat master
master <- build_master(sheets)

# 4. Or enrich just one repeat sheet with selected parent columns
enriched <- enrich_repeat(sheets, "members",
                          parent_cols = c("hh_id", "village"))
```

## Learn More

See the [Getting Started vignette](https://kcham193.github.io/odkmerge/articles/getting-started.html)
for full documentation including multi-repeat and nested repeat examples.

## Code of Conduct

Please note that the `odkmerge` project is released with a
[Contributor Code of Conduct](https://github.com/kcham193/odkmerge/blob/master/.github/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Contributing

Contributions are welcome! See [CONTRIBUTING.md](https://github.com/kcham193/odkmerge/blob/master/.github/CONTRIBUTING.md)
for guidelines. All contributions must pass `devtools::check()` with
0 errors, 0 warnings, and 0 notes.

## License

MIT © Kasim Chambulilo
