---
title: "Getting Started with surveymerge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with surveymerge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE
)
library(surveymerge)
```

## Introduction

XLSForm-based platforms such as **ODK Central** and **KoboToolbox** are
the most widely-used tools for mobile data collection in ecology, public
health, agriculture, and humanitarian research. Their forms support
**repeat groups** - sections of a survey that can be filled in multiple
times per submission. A household survey might repeat for each member;
a vegetation survey might repeat for each step walked along a transect.

When you export your data to Excel, every repeat group lands in its own
sheet, linked back to the parent via index columns. Re-joining those
sheets by hand is repetitive and error-prone - but joining them blindly
into one wide table can be statistically misleading.

`surveymerge` automates the relational plumbing **and** keeps the grain
structure of your data explicit, so you can build analysis-ready
datasets at the unit of analysis you actually care about.

The package handles three structural patterns:

- **Simple**: one parent sheet + one repeat sheet.
- **Multi-repeat**: one parent sheet + multiple sibling repeat sheets.
- **Nested**: a repeat sheet that is itself the parent of another
  repeat sheet.

---

## Understanding units of analysis

Before showing the code, it is worth pausing on the most important idea
in this package: **a survey export is not one dataset, it is several**.

A typical household survey export can contain:

- **Household-level** data - one row per household. Lives in the parent
  sheet (`household`).
- **Member-level** data - one row per member. Lives in a `members`
  repeat sheet.
- **Visit-level** data - one row per visit. Lives in a `visits` repeat
  (if your form has revisits).
- **Observation-level** data - one row per measurement. Lives even
  deeper in nested repeats.

Each of those is a legitimate **unit of analysis**:

| Unit of analysis | Example research question                                   | Where the data lives                     |
|------------------|-------------------------------------------------------------|------------------------------------------|
| Household        | What fraction of households own livestock?                  | parent sheet                             |
| Member           | What is the age distribution of household members?          | `members` repeat (+ selected hh columns) |
| Asset            | Which assets are most common?                               | `assets` repeat                          |
| Visit            | How long between baseline and follow-up visits?             | `visits` repeat                          |
| Observation      | What is the average leaf cover per measurement?             | nested observation repeat                |

A common mistake is to "flatten" all of these into a single wide table.
That looks tidy, but it inflates row counts (a household with five
members is now five rows), distorts means and proportions, and turns
parent-level variables into pseudo-repeated measures. `surveymerge`
gives you one tibble per grain so you can pick the right one.

---

## Installation

```{r install, eval = FALSE}
# From GitHub (development version):
devtools::install_github("kcham193/surveymerge")

# From CRAN (once published):
install.packages("surveymerge")
```

---

## Quick start

For most users, `survey_merge()` is all you need:

```{r quickstart}
path   <- system.file("extdata", "simple_survey.xlsx", package = "surveymerge")
master <- survey_merge(path)
head(master)
```

One function call. One tibble. Aligned to the species grain (one row
per recorded species), with plot-level columns attached.

---

## Understanding the sheet structure

When you export from KoboToolbox or ODK Central, you get an `.xlsx` file
with multiple sheets. The **parent sheet** has one row per submission.
Each **repeat sheet** has one row per repeated entry, and is linked to
its parent via special columns:

| Column                | Sheet               | Meaning                                |
|-----------------------|---------------------|----------------------------------------|
| `_index` / `KEY`      | every sheet         | Row identifier within this sheet       |
| `_parent_index` / `PARENT_KEY` | repeat sheets only | The `_index` value of the parent row   |
| `_parent_table_name`  | repeat sheets only  | Name of the parent sheet (Kobo only)   |

`surveymerge` uses these columns to detect and join sheets automatically,
across both the KoboToolbox and ODK Central export conventions.

---

## Step-by-step workflow

For more control, you can run each step separately.

### Step 1 - Read the export

```{r step1}
path   <- system.file("extdata", "simple_survey.xlsx", package = "surveymerge")
sheets <- read_survey_export(path)
names(sheets)
```

`read_survey_export()` returns a named list of tibbles, one per sheet.
You can inspect any sheet directly:

```{r step1b}
dplyr::glimpse(sheets[["survey"]])
```

### Step 2 - Detect the structure

```{r step2}
structure <- detect_structure(sheets)
print(structure)
```

`detect_structure()` tells you which sheet is the parent, which are
repeats, and how they relate - i.e. which units of analysis your export
contains.

### Step 3 - Build a dataset at the right grain

```{r step3}
master <- build_master(sheets, structure = structure)
dplyr::glimpse(master)
```

Passing the pre-computed `structure` object avoids repeating the
detection step.

---

## Handling multiple repeat groups

Some forms have more than one repeat group at the same level - for
example, a household survey that collects both household members and
household assets as separate repeats.

```{r multi}
path   <- system.file("extdata", "multi_repeat_survey.xlsx", package = "surveymerge")
sheets <- read_survey_export(path)
detect_structure(sheets)
```

When there are multiple repeat sheets, `build_master()` returns a
**named list of tibbles** - one per repeat group. Sibling repeats live
at *different* grains (members vs. assets), so forcing them into a
single frame would conflate units of analysis.

```{r multi2}
result <- build_master(sheets)
names(result)
nrow(result[["members"]])
nrow(result[["assets"]])
```

Each tibble already includes the parent-level columns that make it
self-contained at its grain:

```{r multi3}
colnames(result[["members"]])
```

Use `result[["members"]]` for member-level analysis, `result[["assets"]]`
for asset-level analysis, and the parent sheet directly for
household-level analysis.

---

## Nested repeats

A nested repeat is a repeat group inside another repeat group. In the
example below, a farm has multiple fields, and each field has multiple
observations - three levels deep.

```{r nested}
path   <- system.file("extdata", "nested_survey.xlsx", package = "surveymerge")
sheets <- read_survey_export(path)
detect_structure(sheets)
```

`build_master()` resolves the full chain recursively. The `observation`
master will contain columns from both `field` and `farm`:

```{r nested2}
result <- build_master(sheets)
obs    <- result[["observation"]]
# Confirm grandparent column is present
"farm_id" %in% colnames(obs)
dplyr::glimpse(obs)
```

The observation tibble has one row per observation - the natural grain
for observation-level analysis, with farm- and field-level context
joined in.

---

## Choosing parent columns deliberately

If you only want specific parent columns added to a repeat sheet - not
the whole parent flattened in - use `enrich_repeat()`:

```{r enrich}
path   <- system.file("extdata", "simple_survey.xlsx", package = "surveymerge")
sheets <- read_survey_export(path)

enriched <- enrich_repeat(
  sheets,
  repeat_sheet_name = "species",
  parent_cols       = c("plot_id", "observer", "vegetation_type")
)

colnames(enriched)
```

This is useful when the parent sheet has dozens of columns and you only
need a few covariates for the analysis at the repeat grain. It also
makes your script self-documenting about *which* parent variables you
intentionally carried across.

---

## Migration from `odkmerge`

If you previously used `odkmerge` 0.1.0:

- `odk_merge()` still works but is deprecated; use `survey_merge()`.
- `read_odk_export()` still works but is deprecated; use
  `read_survey_export()`.
- `detect_structure()` now returns an object of class
  `survey_structure` (and still carries the legacy `odk_structure`
  class for back-compat).

Each deprecated call emits a one-time warning per session.

---

## Exporting results

Once you have a dataset at the grain you want, exporting is straightforward:

```{r export, eval = FALSE}
# CSV
write.csv(master, "vegetation_master.csv", row.names = FALSE)

# Excel (single sheet)
writexl::write_xlsx(master, "vegetation_master.xlsx")

# Excel (multiple sheets - for multi-repeat results)
result <- survey_merge(path, verbose = FALSE)
if (is.data.frame(result)) {
  writexl::write_xlsx(result, "master.xlsx")
} else {
  writexl::write_xlsx(result, "master.xlsx")  # named list -> one sheet per repeat
}
```

---

## FAQ

**Q: My form has a repeat but I only see one sheet in the export - why?**

Some deployments only export sheets that have at least one entry. If
all submissions skipped a repeat group, the sheet will be absent. Check
that at least one submission has data in the repeat group, then
re-export.

---

**Q: `survey_merge()` returns a list instead of a single tibble - what
do I do?**

A list is returned when the export contains multiple repeat groups -
because each repeat lives at its own unit of analysis. Pick the tibble
that matches your research question (e.g. `result[["members"]]` for
member-level analysis). Use `detect_structure()` first to see what your
form contains.

---

**Q: Column names are duplicated with `_repeat` and `_parent` suffixes.**

This happens when the same column name exists in both the parent and
repeat sheet (e.g. both have a column called `date`). The suffix makes
the origin explicit. You can change the suffix with the `suffix`
argument: `survey_merge(path, suffix = c("_child", "_main"))`.

---

**Q: Can I use this with CSV exports?**

Currently `surveymerge` supports `.xlsx` exports only. CSV support is
planned for a future version.

---

**Q: `_submission_*` columns are cluttering my output - how do I remove them?**

They are removed by default (`drop_internal = TRUE`). If they are
appearing, check that you are not passing `drop_internal = FALSE`. You
can also remove them after the fact with
`dplyr::select(master, -starts_with("_submission_"))`.

---

**Q: My form has 4 or more levels of nesting - does surveymerge handle that?**

Yes. `build_master()` walks up the ancestry chain recursively, so any
depth of nesting is supported as long as each sheet has valid
`_parent_index` and (on Kobo) `_parent_table_name` columns.
