---
title: "Getting Started with odkmerge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with odkmerge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE
)
library(odkmerge)
```

## Introduction

Open Data Kit (ODK) and KoboToolbox are the most widely-used platforms for
mobile data collection in ecology, public health, agriculture, and humanitarian
research. Both platforms are built on the XLSForm standard, which supports
**repeat groups** — sections of a survey that can be filled in multiple times
per submission. A household survey might repeat for each household member; a
vegetation survey might repeat for each step walked in a transect.

When you export your data to Excel, ODK and KoboToolbox split every repeat group
into its own sheet. A form with one repeat produces a two-sheet workbook; a form
with nested repeats can produce three or more sheets. Each repeat sheet links back
to its parent via two columns: `_index` (in the parent) and `_parent_index` (in
the repeat).

Joining these sheets back together manually is repetitive, error-prone, and has
to be redone every time you re-export your data. `odkmerge` automates the entire
workflow. It detects the sheet structure, resolves the join keys, and returns
flat, analysis-ready tibbles — for any ODK or KoboToolbox export, regardless of
form complexity.

The package handles three structural patterns:

- **Simple**: one parent sheet + one repeat sheet.
- **Multi-repeat**: one parent sheet + multiple sibling repeat sheets.
- **Nested**: a repeat sheet that is itself the parent of another repeat sheet.

---

## Installation

```{r install, eval = FALSE}
# From GitHub (development version):
devtools::install_github("yourusername/odkmerge")

# From CRAN (once published):
install.packages("odkmerge")
```

---

## Quick Start

For most users, `odk_merge()` is all you need:

```{r quickstart}
path   <- system.file("extdata", "simple_survey.xlsx", package = "odkmerge")
master <- odk_merge(path)
head(master)
```

One function call. One flat tibble. Ready to analyse.

---

## Understanding the Sheet Structure

Before diving deeper, it helps to understand what an ODK export looks like.

When you export from KoboToolbox or ODK Central, you get an `.xlsx` file with
multiple sheets. The **parent sheet** (sometimes called the main or survey
sheet) has one row per submission. Each **repeat sheet** has one row per
repeated entry, and contains three special columns:

| Column | Sheet | Meaning |
|--------|-------|---------|
| `_index` | every sheet | Row identifier within this sheet |
| `_parent_index` | repeat sheets only | The `_index` value of the parent row |
| `_parent_table_name` | repeat sheets only | The name of the parent sheet |

`odkmerge` uses these columns to automatically detect and join the sheets.

---

## Step-by-Step Workflow

For more control, you can run each step separately.

### Step 1 — Read the export

```{r step1}
path   <- system.file("extdata", "simple_survey.xlsx", package = "odkmerge")
sheets <- read_odk_export(path)
names(sheets)
```

`read_odk_export()` returns a named list of tibbles, one per sheet. You can
inspect any sheet directly:

```{r step1b}
dplyr::glimpse(sheets[["survey"]])
```

### Step 2 — Detect the structure

```{r step2}
structure <- detect_structure(sheets)
print(structure)
```

`detect_structure()` tells you which sheet is the parent, which are repeats,
and how they relate to each other.

### Step 3 — Build the master dataset

```{r step3}
master <- build_master(sheets, structure = structure)
dplyr::glimpse(master)
```

Passing the pre-computed `structure` object avoids repeating the detection step.

---

## Handling Multiple Repeat Groups

Some forms have more than one repeat group at the same level — for example, a
household survey that collects both household members and household assets as
separate repeats.

```{r multi}
path   <- system.file("extdata", "multi_repeat_survey.xlsx", package = "odkmerge")
sheets <- read_odk_export(path)
detect_structure(sheets)
```

When there are multiple repeat sheets, `build_master()` returns a **named list
of tibbles** — one per repeat group. Sibling repeats have different row
granularity (members vs. assets), so forcing them into a single frame would
not make sense.

```{r multi2}
result <- build_master(sheets)
names(result)
nrow(result[["members"]])
nrow(result[["assets"]])
```

Each tibble already includes all the parent-level columns:

```{r multi3}
colnames(result[["members"]])
```

---

## Nested Repeats

A nested repeat is a repeat group inside another repeat group. In the example
below, a farm can have multiple fields, and each field can have multiple
observations — three levels deep.

```{r nested}
path   <- system.file("extdata", "nested_survey.xlsx", package = "odkmerge")
sheets <- read_odk_export(path)
detect_structure(sheets)
```

`build_master()` resolves the full chain recursively. The `observation` master
will contain columns from both `field` and `farm`:

```{r nested2}
result <- build_master(sheets)
obs    <- result[["observation"]]
# Confirm grandparent column is present
"farm_id" %in% colnames(obs)
dplyr::glimpse(obs)
```

---

## Selecting Parent Columns

If you only want specific parent columns added to a repeat sheet — not all of
them — use `enrich_repeat()` directly:

```{r enrich}
path   <- system.file("extdata", "simple_survey.xlsx", package = "odkmerge")
sheets <- read_odk_export(path)

enriched <- enrich_repeat(
  sheets,
  repeat_sheet_name = "species",
  parent_cols       = c("plot_id", "observer", "vegetation_type")
)

colnames(enriched)
```

This is useful when the parent sheet has dozens of columns and you only need
a few for your analysis.

---

## Exporting Results

Once you have your flat dataset, exporting is straightforward:

```{r export, eval = FALSE}
# CSV
write.csv(master, "vegetation_master.csv", row.names = FALSE)

# Excel (single sheet)
writexl::write_xlsx(master, "vegetation_master.xlsx")

# Excel (multiple sheets — for multi-repeat results)
result <- odk_merge(path, verbose = FALSE)
if (is.data.frame(result)) {
  writexl::write_xlsx(result, "master.xlsx")
} else {
  writexl::write_xlsx(result, "master.xlsx")  # named list -> one sheet per repeat
}
```

---

## FAQ

**Q: My form has a repeat but I only see one sheet in the export — why?**

Some ODK deployments only export sheets that have at least one entry. If all
submissions skipped a repeat group, the sheet will be absent. Check that at
least one submission has data in the repeat group, then re-export.

---

**Q: `odk_merge()` returns a list instead of a single tibble — what do I do?**

A list is returned when there are multiple repeat sheets. Access individual
tibbles with `result[["sheet_name"]]`, or iterate with `purrr::map()`. Use
`detect_structure()` first to see your form's structure.

---

**Q: Column names are duplicated with `_repeat` and `_parent` suffixes.**

This happens when the same column name exists in both the parent and repeat
sheet (e.g. both have a column called `date`). The suffix makes the origin
explicit. You can change the suffix with the `suffix` argument:
`odk_merge(path, suffix = c("_child", "_main"))`.

---

**Q: Can I use this with KoboToolbox API exports (CSV format)?**

Currently `odkmerge` supports `.xlsx` exports only. CSV exports from KoboToolbox
have a slightly different structure. CSV support is planned for a future version.

---

**Q: `_submission_*` columns are cluttering my output — how do I remove them?**

They are removed by default (`drop_internal = TRUE`). If they are appearing,
check that you are not passing `drop_internal = FALSE`. You can also remove them
after the fact with `dplyr::select(master, -starts_with("_submission_"))`.

---

**Q: My form has 4 or more levels of nesting — does odkmerge handle that?**

Yes. `build_master()` walks up the ancestry chain recursively, so any depth of
nesting is supported as long as each sheet has valid `_parent_index` and
`_parent_table_name` columns.
